Notes

### 2022-10-30 appendix.tex
Updates from Overleaf

\subsection{Additive Log ratio transform}
To calculate the determinant of the Jacobian of the inverse transform,
we start by noting that $s = \textrm{exp} \circ \textrm{norm}$, where
$\textrm{exp}$ is the elementwise exponential function and
\textrm{norm} is defined by
\[
  \textrm{norm}(z) = \frac{z}{\textrm{sum}(z) + 1}.
\]
As such, the resulting Jacobian determinant is the product of the
Jacobian determinants of the component functions,
\[
  \absdet{J_s(y)}
  = \absdet{J_{\textrm{exp}}(y)} \absdet{J_{\textrm{norm}}(z)},
\]
where $z = \textrm{exp}(y)$.  The Jacobian for the exponential
function is diagonal, so the determinant is the product of the
diagonal of the Jacobian, which for $y \in \mathbb{R}^{N-1}$ is
\[
  \absdet{J_{\textrm{exp}}(y)} = \textrm{prod}(\exp(y)).
\]
As above, let $z = \exp(y) \in (0, \inf)^{N-1}$.  We can differentiate
$\textrm{norm}$ to derive the Jacobian,
\[
  J_{\textrm{norm}}
  = \frac{1}{1 + \textrm{sum}(z)} \mathbb{I}_{N-1}
  - \left(\frac{1}{(1 + \textrm{sum}(z))^2} \beta \right)
  \textrm{vector}_{N-1}(1)^{\top},
\]
where $\mathbb{I}_{N-1}$ is the $(N - 1) \times (N - 1)$ unit matrix and
$\textrm{vector}_{N-1}(1)$ is the $N - 1$-vector with values 1.  Using
the matrix determinant lemma,\footnote{The matrix determinant lemma
  is \[\textrm{det}(A + u v^{\top}) = (1 + v^{\top} A^{-1} u)
    \textrm{det}(A).\]}
we have
\begin{eqnarray*}
  \textrm{absdet}(J_{\textrm{norm}}(z))
  & = &
  \left(
    1
    + \textrm{vector}_{N-1}(1)^{\top}
    \left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I} \right)^{-1}
    \frac{-z}{(1 + \textrm{sum}(z))^2}
    \right)
    \ \textrm{det}\left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I}
        \right)
  \\[6pt]
  & = &
  \left(
    1 
    + \textrm{sum}\left( \frac{-(1 + \textrm{sum}(z)) z}{(1 +
        \textrm{sum}(z))^2} \right)
  \right)
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = &
        \left(1 + \textrm{sum}\left(\frac{-z}{1 + \textrm{sum}(z)} \right)\right)        
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = & \left( 1 - \textrm{sum}(\textrm{norm}(z)) \right) 
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = & \left( \frac{1}{1 + \textrm{sum}(z)} \right)^N.
\end{eqnarray*}
Thus the entire absolute determinant of the Jacobian is defined by the
product, 
\[
  \absdet{J_s(y)}
  \ = \
  \textrm{prod}(\exp(y))
  \, \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N.
\]
and our final expression for densities for unconstrained $y \in
\mathbb{R}^{N-1}$ is
\[
  p_Y(y)
  = p_X(\textrm{alr}^{-1}(y))
  \, \textrm{prod}(\exp(y))
  \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N
\]  
\end{appendices}

>>>>

\subsection{Additive Log ratio transform}
To calculate the determinant of the Jacobian of the inverse transform,
we start by noting that $s = \textrm{exp} \circ \textrm{norm}$, where
$\textrm{exp}$ is the elementwise exponential function and
\textrm{norm} is defined by
\[
  \textrm{norm}(z) = \frac{z}{\textrm{sum}(z) + 1}.
\]
As such, the resulting Jacobian determinant is the product of the
Jacobian determinants of the component functions,
\[
  \absdet{J_s(y)}
  = \absdet{J_{\textrm{exp}}(y)} \absdet{J_{\textrm{norm}}(z)},
\]
where $z = \textrm{exp}(y)$.  The Jacobian for the exponential
function is diagonal, so the determinant is the product of the
diagonal of the Jacobian, which for $y \in \mathbb{R}^{N-1}$ is
\[
  \absdet{J_{\textrm{exp}}(y)} = \textrm{prod}(\exp(y)).
\]
As above, let $z = \exp(y) \in (0, \inf)^{N-1}$.  We can differentiate
$\textrm{norm}$ to derive the Jacobian,
\[
  J_{\textrm{norm}}
  = \frac{1}{1 + \textrm{sum}(z)} \mathbb{I}_{N-1}
  - \left(\frac{1}{(1 + \textrm{sum}(z))^2} \beta \right)
  \textrm{vector}_{N-1}(1)^{\top},
\]
where $\mathbb{I}_{N-1}$ is the $(N - 1) \times (N - 1)$ unit matrix and
$\textrm{vector}_{N-1}(1)$ is the $N - 1$-vector with values 1.  Using
the matrix determinant lemma,\footnote{The matrix determinant lemma
  is \[\textrm{det}(A + u v^{\top}) = (1 + v^{\top} A^{-1} u)
    \textrm{det}(A).\]}
we have
\begin{eqnarray*}
  \textrm{absdet}(J_{\textrm{norm}}(z))
  & = &
  \left(
    1
    + \textrm{vector}_{N-1}(1)^{\top}
    \left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I} \right)^{-1}
    \frac{-z}{(1 + \textrm{sum}(z))^2}
    \right)
    \ \textrm{det}\left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I}
        \right)
  \\[6pt]
  & = &
  \left(
    1 
    + \textrm{sum}\left( \frac{-(1 + \textrm{sum}(z)) z}{(1 +
        \textrm{sum}(z))^2} \right)
  \right)
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = &
        \left(1 + \textrm{sum}\left(\frac{-z}{1 + \textrm{sum}(z)} \right)\right)        
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = & \left( 1 - \textrm{sum}(\textrm{norm}(z)) \right) 
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = & \left( \frac{1}{1 + \textrm{sum}(z)} \right)^N.
\end{eqnarray*}
Thus the entire absolute determinant of the Jacobian is defined by the
product, 
\[
  \absdet{J_s(y)}
  \ = \
  \textrm{prod}(\exp(y))
  \, \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N.
\]
and our final expression for densities for unconstrained $y \in
\mathbb{R}^{N-1}$ is
\[
  p_Y(y)
  = p_X(\textrm{alr}^{-1}(y))
  \, \textrm{prod}(\exp(y))
  \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N
\]

\subsection{Hyperspherical simplex transforms}

\subsubsection{Hyperspherical coordinate transform}

We use the following definition of hyperspherical coordinates.
For $\phi_i \in (0, \pi/2]$ and $u \in \mathbb{S}^N_{>0}$, the positive orthant of the $N$-sphere,
\[
  u_i = \left(\prod_{k=1}^{i-1} \sin(\phi_k)\right) \begin{cases}
    1 & \text{ if } i = N + 1\\
    \cos(\phi_i) & \text{otherwise}
  \end{cases}.
\]

Let $z_i = \sin^2(\phi_i)$ and $x_i = u_i^2$, so that $x \in \Delta^N$.
Then
\[
  x_i = \left(\prod_{k=1}^{i-1} z_k\right) \left(\begin{cases}
    1 & \text{ if } i = N + 1\\
    (1 - z_i) & \text{otherwise}
  \end{cases}\right) = (1 - z_i (1 - \delta_{i,N+1})) \prod_{k=1}^{i-1} z_k.
\]

\subsubsection{Hyperspherical coordinates inverse transform}

To derive the inverse transform, note that
\[
    \begin{aligned}
        x_{N+1} &= \prod_{k=1}^{N} z_k\\
        x_N + x_{N+1} &= (1 - z_N)\left(\prod_{k=1}^{N-1} z_k\right) + z_N \left( \prod_{k=1}^{N-1} z_k \right) = \prod_{k=1}^{N-1} z_k = \frac{x_N}{1 - z_N}\\
        x_{N-1} + x_{N} + x_{N+1} &= (1-z_{N-1}) \left(\prod_{k=1}^{N-2} z_k\right) + z_{N-1} \left(\prod_{k=1}^{N-2} z_k\right) = \prod_{k=1}^{N-2} z_k = \frac{x_{N-1}}{1 - z_{N-1}}\\
        &\vdots \\
        \sum_{k=i}^{N+1} x_{k} &= \prod_{k=1}^{i - 1} z_k = \frac{x_i}{1 - z_i}.
    \end{aligned}
\]

From the final expression, we can define the inverse transform:
\[
z_i = 1 - \frac{x_i}{\sum_{k=i}^{N+1} x_k}
\]

\subsubsection{Jacobian of hyperspherical coordinates transform}

To derive the Jacobian, we note that $x_i$ depends only on $z_k$ for $k \le i$.
As a result, the Jacobian of the map from $z$ to $x_-$ is lower triangular, and its determinant depends only on its diagonal, which consists of the terms
$$J_{ii} = -\frac{\partial x_i}{\partial z_i} = \prod_{k=1}^{i-1} z_k,$$
so
$$
|J| = \prod_{i=1}^N |J_{ii}|
= \prod_{i=1}^N \prod_{k=1}^{i-1} z_k
= \prod_{k=1}^{N-1} \prod_{i=k+1}^N z_k
= \prod_{i=1}^{N-1} z_i^{N-i} = \prod_{i=1}^N z_i^{N-i}
$$

Due to the form of the Jacobian correction, if $x$ is uniformly distributed on the unit simplex, then $z_i$ and $z_j$ are uncorrelated for all $i \ne j$.

\subsubsection{Hyperspherical angular transform}

The map from $y$ to $z$ is elementwise, so its Jacobian is diagonal.
Each element is
$$\sin^{-1}(\sqrt{z_i}) = \phi_i = \frac{\pi}{2} \operatorname{logit}^{-1} (y_i).$$
Differentiating all sides with respect to $y_i$, we find
$$
\begin{aligned}
\frac{1}{\sqrt{1 - z_i^2}} \frac{1}{2 \sqrt{z_i}} \frac{\partial z_i}{\partial y_i} &= \frac{\pi}{2} \operatorname{logit}^{-1} (y_i) (1 - \operatorname{logit}^{-1} (y_i)) = \phi_i \left(1 - \frac{2}{\pi} \phi_i\right)\\
\frac{\partial z_i}{\partial y_i} &= 2 \phi_i \left(1 - \frac{2}{\pi} \phi_i\right) \sin(\phi_i) \cos(\phi_i).
\end{aligned}
$$

As a result, the combined transform $f: y \mapsto x$ is
\[
  x_i = \left(1 - \cos^2\left(\frac{\pi}{2} \operatorname{logit}^{-1} (y_i)\right) (1 - \delta_{i,N+1})\right)\left(\prod_{k=1}^{i-1} \sin^2\left(\frac{\pi}{2} \operatorname{logit}^{-1} (y_k)\right)\right),
\]
and its inverse is
\[
z_i = \operatorname{logit}\left(\frac{2}{\pi}\sin^{-1}\left(\sqrt{1 - \frac{x_i}{\sum_{k=i}^{N+1} x_k}}\right)\right).
\]

The Jacobian determinant is
$$
\begin{aligned}
    |J_f| &= \prod_{i=1}^N z_i^{N-i} 2 \phi_i \left(1 - \frac{2}{\pi} \phi_i\right) \sin(\phi_i) \cos(\phi_i)\\
          &= 2^N \prod_{i=1}^N \phi_i \left(1 - \frac{2}{\pi} \phi_i\right) \sin^{2(N-i)+1}(\phi_i) \cos(\phi_i).
\end{aligned}
$$

\subsubsection{Hyperspherical logit transform}

For $z_i \in (0, 1]$, $z_i^k \in (0, 1]$ is bijective for $k \ne 0$.
Let $g(w)$ be the density function for a distribution on $\mathbb{R}$ and $G(w)$ be its cumulative distribution function.
By the definition of the CDF, $\frac{\mathrm{d}}{\mathrm{d}w} G(w) = g(w)$.
Let $z_i^{N-i+1} = G(y_i)$.

The combined transform $f: y \mapsto x$ is then
\[
  x_i = \left(1-G(y_i)^{1/(N-i+1)}(1 - \delta_{i,N+1})\right)\left(\prod_{k=1}^{i-1} G(y_k)^{1/(N-k+1)}\right),
\]
and its inverse is 
\[
y_i = G^{-1}\left(\left(1 - \frac{x_i}{\sum_{k=i}^{N+1} x_k}\right)^{N-i+1}\right).
\]

Then
$$
\begin{aligned}
(N-i+1) z_i^{N-i} \frac{\partial z_i}{\partial y_i} &= g(y_i)\\
\frac{\partial z_i}{\partial y_i} &= \frac{g(y_i)}{(N-i+1) z_i}.
\end{aligned}
$$

As a result,
$$
\begin{aligned}
|J_f| &= \prod_{i=1}^{N} z_i^{N-i} \frac{g(y_i)}{(N-i+1) z_i^{N-i}}\\ 
&= \left(\prod_{i=1}^N \frac{1}{N-i+1} \right) \prod_{i=1}^{N} g(y_i)\\
&= \frac{1}{N!} \prod_{i=1}^{N} g(y_i).
\end{aligned}
$$

This particular transform allows us to transform the uniform distribution on the simplex to any distribution on $\mathbb{R}^N$ whose coordinates are IID distributed according to $g$.

If we select $G$ to be the logistic function, then $g(w) = \operatorname{logit}^{-1}(w) (1 - \operatorname{logit}^{-1}(w))$ is the density of the logistic distribution with mean $0$ and scale term $1$, and $y_i$ is logistically distributed.
This choice of $G$ defines the hyperspherical logit transform.

\subsubsection{Hyperspherical probit transform}

Similarly, to define the hyperspherical probit transform, we select $G$ to be the inverse probit function.
Then $g$ is the density function of the standard normal distribution.

\subsubsection{Transforming the Dirichlet distribution}

Recall that the Dirichlet distribution has density $p_X(x)$ equal to
\[
p_X(x) \propto \prod_{i=1}^{N+1} x_i^{\alpha_i - 1}.
\]

Note that
\[
\begin{aligned}
\prod_{i=1}^{N+1} x_i^{\alpha_i - 1} &= \prod_{i=1}^{N+1} \left(\left(1-G(y_i)^{1/(N-i+1)}(1 - \delta_{i,N+1})\right)\left(\prod_{k=1}^{i-1} G(y_k)^{1/(N-k+1)}\right)\right)^{\alpha_i - 1}\\
&= \left(\prod_{i=1}^{N} \left(1-G(y_i)^{1/(N-i+1)}\right)^{\alpha_i - 1}\right) \prod_{i=1}^{N+1}\prod_{k=1}^{i-1} G(y_k)^{(\alpha_i - 1)/(N-k+1)}\\
&= \prod_{i=1}^{N} \left(1-G(y_i)^{1/(N-i+1)}\right)^{\alpha_i - 1} G(y_i)^{-1+\sum_{k=i+1}^{N+1}\alpha_k/(N-i+1)}\\
\end{aligned}
\]
The transformed density on $y$ is then
\[
p_Y(y) \propto \prod_{i=1}^{N} \left(1-G(y_i)^{1/(N-i+1)}\right)^{\alpha_i - 1} G(y_i)^{-1+\sum_{k=i+1}^{N+1}\alpha_k/(N-i+1)} g(y_i) = \prod_{i=1}^{N} h(y_i, \alpha_{i:N+1}),
\]
where $h$ is a density function.
Therefore, regardless of the choice of $g$ or $\alpha$, the individual coordinates of the transformed distribution are independently distributed.
\end{appendices}

### 2022-10-30 notes.tex
Updates from Overleaf

The correction term $p_Y(y) = p_X(f^{-1}(y))\,
\prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right).$
\subsection{Additive log ratio transform}

@@

@@

$\Exp(Y)$ is one example of a matrix function $f$, which we can roughly define for symmetric matrices as any function that obeys
\[f(Y) = U f(\Lambda) U^\top,\]
where $f(\Lambda)_{ij} = f(\lambda_i)\delta_{ij}$.
\footnote{
  In a similar way, we can generalize other scalar functions like the inverse, trigonometric, and hyperbolic functions to functions of matrices.
  All results in this section apply to these functions as well.
}

For such functions $f$, we can compute a common form of the Jacobian determinant.
Differentiating the expression, we find
\[
\begin{aligned}
  \dv{(f(Y))} &= \dv{U} f(\Lambda) U^\top + U f'(\Lambda) \dv{\Lambda} U^\top + U f(\Lambda) \dv{U}^\top\\
  U^\top \dv{(f(Y))} U &= U^\top \dv{U} f(\Lambda) + f(\Lambda) \dv{U}^\top U + f'(\Lambda)\dv{\Lambda},
\end{aligned}
\]
where $f'(\Lambda)_{ij} = \frac{\dv{(f(\lambda_i))}}{\dv{\lambda_i}}$.

>>>>

The correction term $p_Y(y) = p_X(f^{-1}(y))\,
\prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right).$

\subsection{Additive log ratio transform}

@@

\subsection{Isometric Log Ratio Transform}

We may define a distance preserving transformation from $\textrm{ilr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ called an isometric log ratio transform. By noting that the ALR transformation may be written as

\[
  \begin{aligned}
 \textrm{alr}(x) &= \begin{bmatrix}\displaystyle
    \log \frac{x_1}{x_N} \cdots \log \frac{x_{N-1}}{x_N}
  \end{bmatrix} \\
  &= \log(\mathbf{x}) \cdot \begin{bmatrix}\displaystyle
                    1 & 0 & \cdots & 0 \\
                    0 & 1 & \cdots & 0 \\
                    \vdots & \vdots & \ddots & \vdots \\
                    0 & 0 & \cdots & 1 \\
                    -1 & -1 & \cdots & -1
                    \end{bmatrix}.
  \end{aligned}
\]
Take any unitary matrix $\mathbf{V} \in D \times D - 1$ then the ILR is defined for $x \in \Delta^{N-1}$ by
\[
  \textrm{ilr}(x)
  = \log(\mathbf{x}) \cdot \mathbf{V}.
\]

Unitary $\mathbf{V}$ matrices may be constructed through the Gram-Schmidt process or by normalizing the rows of a Helmert contrast matrix using the L2 norm.

The inverse isometric log ratio transform maps values in
$\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
\mathbb{R}^{N-1}$ by
\[
  \textrm{ilr}^{-1}(y)
  = \textrm{softmax}(\begin{bmatrix} \exp(\mathbf{V^{-1}} \mathbf{x} ) \end{bmatrix}),
\]
where for $u \in \mathbb{R}^N$,
\[
  \textrm{softmax}(u) = \frac{\exp(u)}{\textrm{sum}(\exp(u))}.
\]

The affine transformation of the multiplication by the inverse unitary matrix does not change
the calculation of the change of variables adjustment proceeds similarly to the ALR case. In fact, any affine transformation may be completed at this stage which generalizes the ILR to power transformations as in \cite{tsagris2011data}\footnote{
   Let $\alpha$ be the scalar power transformation parameter in $\mathbb{R}$ then the inverse \textit{power} based transformation is given by $(V^{-1} \mathbf{x} \alpha + 1) / \alpha$. See \cite{tsagris2011data} for more information.
}.

Given a density $p_X(x)$ defined over simplices $x \in \Delta^{N-1}$,
we can transform to a density over unconstrained parameters $y \in
\mathbb{R}^{N-1}$ by applying the inverse ILR transform and adjusting
for the change of variables, which yields
\[
  p_Y(y) = p_X(\textrm{ilr}^{-1}(y)) \absdet{J_{s}(y)},
\]
where $J_{s}(y)$ is the Jacobian of the function $s$ evaluated at $y$
and $\absdet{J_s(y)}$ is the absolute value of its determinant.

@@

$\Exp(Y)$ is one example of a matrix function $f$, which we can roughly define for symmetric matrices as any function that obeys
\[f(Y) = U f(\Lambda) U^\top,\]
where $f(\Lambda)_{ij} = f(\lambda_i)\delta_{ij}$.
\footnote{
  In a similar way, we can generalize other scalar functions like the inverse, trigonometric, and hyperbolic functions to functions of matrices.
  All results in this section apply to these functions as well.
}
For such functions $f$, we can compute a common form of the Jacobian determinant.
Differentiating the expression, we find
\[
\begin{aligned}
  \dv{(f(Y))} &= \dv{U} f(\Lambda) U^\top + U f'(\Lambda) \dv{\Lambda} U^\top + U f(\Lambda) \dv{U}^\top\\
  U^\top \dv{(f(Y))} U &= U^\top \dv{U} f(\Lambda) + f(\Lambda) \dv{U}^\top U + f'(\Lambda)\dv{\Lambda},
\end{aligned}
\]
where $f'(\Lambda)_{ij} = \frac{\dv{(f(\lambda_i))}}{\dv{\lambda_i}}$.
